---
title: "PROJECT 1 REPORT"
author:
- Itunu Olufemi
- Abdullah Akinde
- Idris Ajibade
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: yes
    toc: yes
    toc_depth: 4
  word_document:
    toc: yes
    toc_depth: '4'
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{amsfonts}
- \usepackage{subfig}
- \usepackage{graphicx}
- \usepackage{amsthm}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{MATH 5310 -- Machine Learning}
- \lhead{Project I}
- \cfoot{\thepage}
- \usepackage{algorithm}
- \usepackage[noend]{algpseudocode}
geometry: margin=1in
spacing: single
fontsize: 11pt
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


\section{Problem 1 Data Scraping}

For this problem, we used the downloaded data set from D2L as the URL method returned an "HTTP status was 403 Forbidden" error.
```{r, message=F, warning=F}
#load in the data set
data1 <- read.csv("~/APSUGradSchool/_Fall 2021/MATH 5310/week 4/_Project 1/weather-2021.csv", 
                 check.names = F)

dim(data1)
```
The dimension of the data set is 255 rows by 22 columns.

```{r, message=F, warning=F}
head(data1) 
```
The above shows the first 6 rows of the data set.

```{r, message=F, warning=F}
tail(data1) 
```
The above represents the last 6 rows of the data set.

\section{Problem 2 Data Cleaning and Preparation}

## Problem 2A remove the "Time of maximum wind gust" in the 10th column
```{r, message=F, warning=F}

#remove a second columns of all the rows  [r x c]
data1 <- data1[, -c(10)]

dim(data1)

```
We have successfully removed the 10th column, our new dimension is now 255 by 21.

## Problem 2B rename the dataset
```{r, message=F, warning=F}

names(data1) <- c("Month", "Date", "MinTemp", "MaxTemp", "Rainfall",
                "Evaporation", "Sunshine", "WindGustDir", "WindGustSpeed",
                "Temp9am", "Humidity9am", "Cloud9am", "WindDir9am",
                "WindSpeed9am", "Pressure9am", "Temp3pm", "Humidity3pm",
                "Cloud3pm", "WindDir3pm", "WindSpeed3pm", "Pressure3pm")

```

## Problem 2C print out unique values
```{r, message=F, warning=F}

data <- data1
vnames <- colnames(data)
n <- nrow(data)
out <- NULL
for (j in 1:ncol(data)){
  vname <- colnames(data)[j]
  x <- as.vector(data[,j])
  n1 <- sum(is.na(x), na.rm=TRUE)  # NA
  n2 <- sum(x=="NA", na.rm=TRUE) # "NA"
  n3 <- sum(x==" ", na.rm=TRUE)  # missing
  nmiss <- n1 + n2 + n3
  nmiss <- sum(is.na(x))
  ncomplete <- n-nmiss
  out <- rbind(out, c(col.num=j, v.name=vname, mode=mode(x), n.level=length(unique(x)),
                      ncom=ncomplete, nmiss= nmiss, miss.prop=nmiss/n))
}
out <- as.data.frame(out) 
row.names(out) <- NULL
out

```
'
There are 21 variables with their respective datatypes for 255 values. This includes:

 1. Month       : which consists of character datatype and have 9 unique values.
 2. Date        : which consists of character datatype and have 255 unique values. This is because each date is unique
 3. MinTemp     : which consists of numeric datatype and have 152 unique values.
 4. MaxTemp     : which consists of numeric datatype and have 156 unique values.
 5. Rainfall    : which consists of numeric datatype and have 46 unique values.
 6. Evaporation : which consists of logical datatype and have 1 unique values.
 7. Sunshine    : which consists of logical datatype and have 1 unique values.
 8. WindGustDir : which consists of character datatype and have 17 unique values.
 9. WindGustSpeed: which consists of integer datatype and have 33 unique values
 10. Temp9am      : which consists of numeric datatype and have 146 unique values
 11. Humidity9am  : which consists of integer datatype and have 56 unique values
 12. Cloud9am     : which consists of integer datatype and have 9 unique values
 13. WindDir9am   : which consists of character datatype and have 17 unique values.
 14. WindSpeed9am : which consists of character datatype and have 23 unique values.This should be integer
 15. Pressure9am  : which consists of numeric datatype and have 173 unique values
 16. Temp3pm      : which consists of numeric datatype and have 152 unique values
 17. Humidity3pm  : which consists of integer datatype and have 74 unique values
 18. Cloud3pm     : which consists of integer datatype and have 9 unique values
 19. WindDir3pm   : which consists of character datatype and have 17 unique values
 20. WindSpeed3pm : which consists of character datatype and have 19 unique values. This should be integer
 21. Pressure3pm  : which consists of numeric datatype and have 174 unique values

## Problem 2D Show the frequency table of each value, as if categorical
```{r, message=F, warning=F}
#use apply column by column
data1 <- data
apply(data1, 2, FUN = function(X){table(X, useNA="ifany")})
```
We identified 1 suspicious variables apart from "WindSpeed9am". This was "WindSpeed3pm" which also had 3 entries with name "Calm"

## Problem 2E change "Calm" to 0, then format to numeric
```{r, message=F, warning=F}
#fixing Windspeed9am 
data1[data1$WindSpeed9am == "Calm", ]$WindSpeed9am <- 0
data1$WindSpeed9am <- as.numeric(data1$WindSpeed9am)
```

```{r, message=F, warning=F}
data1$WindSpeed9am
```


```{r, message=F, warning=F}
#fixing Windspeed3pm

data1[!is.na(data1$WindSpeed3pm == "Calm"), ]$WindSpeed3pm <- 0
data1$WindSpeed3pm <- as.numeric(data1$WindSpeed3pm)
```
```{r, message=F, warning=F}
data1$WindSpeed3pm 
```

```{r, message=F, warning=F}
data1$RainToday <- ifelse(data1$Rainfall > 1, 1, 0)
data1$RainTomorrow <- c(data1$RainToday[2:nrow(data1)], NA)
```

## Problem 2F saving the cleaned dataset
```{r, message=F, warning=F}
#save to csv 
#?write.csv()
write.csv(data1, file="C:\\Users\\john\\Documents\\APSUGradSchool\\_Fall 2021\\MATH 5310\\_project 1\\cleaned-data.csv", row.names =FALSE)
```

\section{Problem 3 Exploratory Data Analysis}
```{r, message=F, warning=F}
set.seed(500) #reproducibility
tab <- table(data1$Month, data1$WindGustDir, useNA="no");
tab

```
We noticed some strong NNW and NW Winds for June, July and Aug. This means some form of association. 

Null Hypothesis H0: is that there is no association between Month and WindGustDir.
Alt Hypothesis H1: is that there is a association between Month and WindGustDir.

Assume our confidence level, alpha to be 0.01
```{r, message=F, warning=F}
fisher.test(tab, simulate.p.value =TRUE)

```
Therefore since p-value is less than 0.1, we reject the Null Hypothesis HO and conclude that there is some form of association between Months and the WindGustDir.

```{r, message=F, warning=F}
set.seed(500) #reproducibility
tab <- table(data1$Month, data1$WindGustSpeed, useNA="no");
tab

```
We noticed no form of association between Month and WindGustSpeed. 

Null Hypothesis H0: is that there is no association between Month and WindGustSpeed.
Alt Hypothesis H1: is that there is a association between Month and WindGustSpeed.

Assume our confidence level, alpha to be 0.01
```{r, message=F, warning=F}
chisq.test(tab, simulate.p.value =TRUE)

```

Therefore since p-value is greater than 0.1, we accept the Null Hypothesis HO and conclude that there is no form of association between Months and the WindGustSpeed.

```{r, message=F, warning=F}
set.seed(500) #reproducibility
hist(data1$MinTemp, useNA="no")

```

We noticed the histogram of Minimum Temperature does not follow a normal distribution 

Null Hypothesis H0: is that there is no association between RainToday and MinTemp.
Alt Hypothesis H1: is that there is a association between RainToday and MinTemp.

Assume our confidence level, alpha to be 0.00001
```{r, message=F, warning=F}
wilcox.test(data1$MinTemp~data1$RainToday, data=data1, alternative = "two.sided")
#chisq.test(tab, simulate.p.value =TRUE)

```

Therefore since p-value is greater than 0.00001, we accept the Null Hypothesis HO and conclude that there is no association between RainToday and MinTemp.

```{r, message=F, warning=F}
set.seed(500) #reproducibility
hist(data1$Pressure3pm, useNA="no")

```
We noticed the histogram of Minimum Temperature nearly follow a normal distribution 

Null Hypothesis H0: is that there is no association between RainToday and Pressure at 3pm.
Alt Hypothesis H1: is that there is a association between RainToday and Pressure at 3pm.

Assume our confidence level, alpha to be 0.01
```{r, message=F, warning=F}
t.test(data1$Pressure3pm~data1$RainToday, data=data1, alternative = "two.sided")


```
Therefore since p-value is lesser than 0.00001, we reject the Null Hypothesis HO and conclude that there is some form of association between RainToday and Pressure at 3pm.